Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-02 paper.tex

\subsection{Tree inference from HMM sample clusters}
\label{sec-sample-cluster-tree-inference}
For a given daily batch of sample sequences that are to be attached to the
ARG, the LS HMM provides the most likely copying path and required mutations
for each of those sequences. We could directly attach these samples to
the ARG according to the copying paths and sets of mutations, but
such an approach would be highly unparsimonious because there are
often many samples that share the same copying path, with
complex patterns of shared mutations among those clusters of samples.
In this section we discuss the strategy used to infer likely tree
topologies within these clusters of samples sharing the same copying path.
In the majority of cases the copying path will consist of a single
node (i.e., no recombination), and so for simplicity we will assume in the
following discussion that all of the samples in a HMM-defined cluster
attach to a single node. The special treatment afforded to
newly-detected putative recombinants are discussed below
in Section XXX.

@@

\subsection{Parsimony increasing hueristics}
Attaching trees built from the clusters of samples that copy from
a particular node (or path of nodes for recombinants, see the
next section) under the HMM is an inherently greedy strategy,
and can produce inferences that are clearly unparsimonious.
The final step in adding a daily batch of samples to the ARG
is then to perform some local updates that target specific
types of parsimony violation in the just-updated regions of the
ARG. There are currently two parsimony-increasing operations
applied, which we refer to as ``mutation collapsing'' and ``reversion
pushing''.

@@

The LS HMM may infer identical paths and patterns of mutations
for multiple samples in a daily batch, and so we
create a ``recombination`` node (marked with a specific ``flags'' value)
for each distinct recombinant. (This node is not strictly necessary
but makes it convenient to find recombinants for subsequent analysis.)
Variation within a cluster
of recombinant sequences is handled in the same way as non-recombinants
(see Section~\ref{sec-sample-cluster-tree-inference} above for details).

@@

\subsection{Node dating}
[Discuss how we date nodes, and suggest caution on interpretation]

% A recombinant sequence was deemed to be HMM-consistent if (1) the same number
% of parent nodes and (2) the same set of mutations immediately ancestral to it
% were obtained from both the unmirrored and mirrored HMM runs. In the ideal
% case, the copying paths of both the HMM runs switch between parent sequences at
% the same position(s); in less ideal cases, however, the copying paths switch at
% different positions, typically because of a lack of genetic variation between
% the positions. The switch positions in the copying paths of both the HMM runs
% were used to define the breakpoint interval(s) of an HMM-consistent
% recombinant.

\subsection{Data preprocessing}

The findings of this study are
based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
up to Aug 22, 2022 and accessible at
% FIXME
doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
if they (1) had ambiguous collection dates; (2) were collected before January
01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
inference, sequences were excluded if they had a ``bad'' quality control status
in any of the four Nextclade columns (``qc.missingData.status'',
``qc.mixedSites.status'', ``qc.frameShifts.status'' and
`qc.stopCodons.status''). These excluded genomes are enriched for low-coverage,
contaminated, or misassembled sequences.

Ambiguous nucleotide letters (i.e.,
not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
Problematic bases in the alignments, which had two or more ambiguous letters or
gaps within a distance of seven bases, were masked as missing data following
the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases
were ignored when attaching a sample genome to the ARG, but they were
subsequently imputed. Further, 481 problematic sites flagged as prone to
sequencing errors or highly homoplasic were excluded entirely
% Do we need this first URL?
(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

\subsection{Exclusion of ``time travellers''}
[FIXME I'm pretty sure we don't have two filters here, we just use
maximum sampling delay. Also not sure the MSD acronym is worth it,
we don't use it much.]

>>>>

\subsection{Tree inference from HMM sample clusters}
\label{sec:sample-cluster-tree-inference}
For a given daily batch of sample sequences that are to be attached to the
ARG, the LS HMM provides the most likely copying path and required mutations
for each of those sequences. We could directly attach these samples to
the ARG according to the copying paths and sets of mutations, but
such an approach would be highly unparsimonious because there are
often many samples that share the same copying path, with
complex patterns of shared mutations among those clusters of samples.
In this section we discuss the strategy used to infer likely tree
topologies within these clusters of samples sharing the same copying path.
In the majority of cases the copying path will consist of a single
node (i.e., no recombination), and so for simplicity we will assume in the
following discussion that all of the samples in a HMM-defined cluster
attach to a single node. The special treatment afforded to
newly-detected putative recombinants are discussed below
in Section XXX.

@@

\subsection{Parsimony increasing hueristics}
\label{sec:parsimony-heuristics}
Attaching trees built from the clusters of samples that copy from
a particular node (or path of nodes for recombinants, see the
next section) under the HMM is an inherently greedy strategy,
and can produce inferences that are clearly unparsimonious.
The final step in adding a daily batch of samples to the ARG
is then to perform some local updates that target specific
types of parsimony violation in the just-updated regions of the
ARG. There are currently two parsimony-increasing operations
applied, which we refer to as ``mutation collapsing'' and ``reversion
pushing''.

@@

The LS HMM may infer identical paths and patterns of mutations
for multiple samples in a daily batch, and so we
create a ``recombination`` node (marked with a specific ``flags'' value)
for each distinct recombinant. (This node is not strictly necessary
but makes it convenient to find recombinants for subsequent analysis.)
Variation within a cluster
of recombinant sequences is handled in the same way as non-recombinants
(see Section~\ref{sec:sample-cluster-tree-inference} above for details).

@@

\subsection{Node dating}
The approaches to assigning a date to nodes in \texttt{sc2ts} is currently
ad-hoc, and the inferred timing of events from the ARGs reported
here should be treated with caution (e.g., Figure~\ref{fig:recomb_mrcas}).
Sample nodes (those corresponding to observed sample sequences) are the
most accurately dated, as we use the reported collection date for
these nodes. These are not entirely accurate, but our data filtering
criteria should remove the most egregious errors (see
Section~\ref{sec:filtering_time_travellers}). Other nodes in the ARG
are dated by splitting the time between the attached samples
and the chosen parent nodes equally (in the case of
trees inferred from daily sample clusters,
Section~\ref{sec:sample-cluster-tree-inference}) or by adding arbitrary
small values when creating new nodes using parsimony rules
(Section~\ref{sec:parsimony-heuristics}). Substantial improvements could
likely be made using simple node dating methods.

\subsection{Data preprocessing}
The findings of this study are
based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
up to Aug 22, 2022 and accessible at
\url{https://doi.org/10.55876/gis8.230329cd} (Supplementary Table 1).
We removed sequences
if they (1) had ambiguous collection dates;
(2) were collected before January 01, 2020 or after June 30, 2022; or
(3) were isolated from a non-human host.
We aligned sequences to the Wuhan-Hu-1/2019 reference sequence
(GenBank: MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
inference, we also excluded sequences if they had a ``bad'' quality control status
in any of the four Nextclade columns (``qc.missingData.status'',
``qc.mixedSites.status'', ``qc.frameShifts.status'' and `qc.stopCodons.status'').

% FIXME this isn't quite right
We encode ambiguous nucleotide letters (i.e.,
not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
Problematic bases in the alignments, which had two or more ambiguous letters or
gaps within a distance of seven bases, were masked as missing data following
the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases
were ignored when attaching a sample genome to the ARG, but they were
subsequently imputed. Further, 481 problematic sites flagged as prone to
sequencing errors or highly homoplasic were excluded entirely
% Do we need this first URL?
(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

\subsection{Filtering ``time travellers''}
\label{sec:filtering_time_travellers}
[FIXME I'm pretty sure we don't have two filters here, we just use
maximum sampling delay. Also not sure the MSD acronym is worth it,
we don't use it much.]
